<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{R/simcross User Guide}
-->

# R/simcross User Guide

[R/simcross](https://github.com/kbroman/simcross) is an
[R](http://www.r-project.org) package to simulate genotypes from an
experimental cross. The aim is flexibility rather than speed.

Meiosis is simulated following the Stahl model (Copenhaver et
al. 2002) with the interference parameter being an integer. In this
model, chiasmata locations are the superposition of two processes: a
proportion _p_ come from a process exhibiting no interference, and the
remainder (proportion 1-_p_) come from a process following the
chi-square model (Foss et al. 1993; Zhao et al. 1995). Thus, with
_p_=0, the model reduces to the chi-square model.  The chi-square
model has a single parameter, _m_, which is a non-negative integer and
controls the strength of interference. _m_=0 corresponds to no
interference.  Broman et al. (2002) estimated _m_=10 for the level of
interference in the mouse (derived assuming _p_=0).

In many organisms, there is always at least one chiasmata on the
four-strand bundle at meiosis. Simulations with
[R/simcross](https://github.com/kbroman/simcross) may be performed
with this assumption; the same model is used, but with rejection
sampling to get results conditional on there being at least one
chiasma. Note that this isn't an assumption of an obligate
_crossover_; if there is exactly one chiasma on the four-strand
bundle, then a random meiotic product will have 0 or 1 crossover, with
probability 1/2 each. Also note that, with the obligate chiasma
assumption, chromosomes must be greater than 50 cM. And note that

## Basic functions

There are two basic functions for simulating a cross:
`create_parent` and `cross`.

`create_parent` parent is for generating a parent object, either an
inbred individual or the F<sub>1</sub> offspring of two inbred
individuals. It takes two arguments: the length of the chromosome in
cM, and the allele or pair of alleles.

Here's how to generate two inbred individuals, one from each of strain
1 and strain 2, and an F<sub>1</sub> individual.

```{r set.seed, echo=FALSE, results="hide"}
set.seed(80605735)
```

```{r create_parent}
library(simcross)
p1 <- create_parent(100, 1)
p2 <- create_parent(100, 2)
f1 <- create_parent(100, 1:2)
```

The `cross` function is used to generate one progeny from the cross
between two individuals.  The input is a pair of individuals (e.g., as
produced by `create_parent`), the interference parameter `m` (default
`m=10`), the parameter `p` for the Stahl model (default `p=0`), `xchr`
to indicate whether the chromosome being simulated is the X chromosome
(`FALSE`, the default, to simulate an autosome), and `male` to
indicate whether the offspring is to be male (which only matters if
`xchr=TRUE`).

Here's an example, to generate an F<sub>2</sub> individual.

```{r sim_f2}
f2 <- cross(f1, f1)
print(f2)
```

The output is list with two components: the maternal and paternal
chromosomes. Each chromsome is a data frame with the allele in a set
of intervals, and the locations of the right endpoints of the
intervals.

In the example above, the maternal chromosome is a non-recombinant `1`
chromosome. The paternal chromosome has two crossovers, with the
the `2` allele up to `r round(f2$pat$locations[1], 1)` cM, the `1` allele
in the interval `r round(f2$pat$locations[1], 1)` &ndash;
`r round(f2$pat$locations[2], 1)`, and the `2` allele
for the remainder of the chromosome.

There's also a function `sim_meiosis` for simulating a meiotic product
from a single individual; the `cross` function calls `sim_meiosis` twice
(once for each parent) and then combines the results into a single
individual.

## Generating pedigrees

The functions `sim_ril_pedigree`, `sim_ail_pedigree`, and
`sim_do_pedigree`.

## Simulating from a pedigree

The function `sim_from_pedigree`.

## Generating marker data

The function `convert2geno` and `get_geno`.




## References

Broman KW, Rowe LB, Churchill GA, Paigen K (2002)
[Crossover interference in the mouse](http://www.genetics.org/content/160/3/1123.short).
Genetics 160:1123-1131

Copenhaver, G. P., Housworth, E. A. and Stahl, F. W. (2002)
[Crossover interference in arabidopsis](http://www.genetics.org/content/160/4/1631.short).
Genetics 160:1631-1639.

Foss, E., Lande, R., Stahl, F. W. and Steinberg, C. M. (1993)
[Chiasma interference as a function of genetic distance](http://www.genetics.org/content/133/3/681.short).
Genetics 133:681-691.

Zhao H, Speed TP, McPeek MS (1995)
[Statistical analysis of crossover interference using the chi-square model](http://www.genetics.org/content/139/2/1045.short).
Genetics 139:1045-1056

<!-- the following to make it look nicer -->
<link href="http://kbroman.org/qtlcharts/assets/vignettes/vignette.css" rel="stylesheet"></link>
